The promise of community driven preprints in ecology and evolution

### 2024-04-19 ms/ms.qmd
Updated paper

Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) encouraging constructive and critical feedback from a research community; 3) helping establish precedence of research findings; and 4) removing financial barriers to publication [e.g., @Proulx2013]. *EcoEvoRxiv* is a community driven preprint server that has paved the way for new initiatives, including accepting multilingual preprints, recognising and accepting registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI).

Preprint servers, such as *EcoEvoRxiv*, can empower researchers to make their findings more accessible, open and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives and allow us to begin to break down barriers to their use in our field.

Here, we explore the first preprints uploaded to *EcoEvoRxiv*, one of the few community run preprint servers, to understand preprint practices and use in our field, focusing our attention on understanding: 1) how long it takes for a preprint to become published, and how many preprints remain unpublished; 2) what countries make use of preprints; 3) how career stage and gender impact preprint use; 4) whether data and code are more likely to be shared in preprints; and 5) the extent to which authors make use of preprint servers for registered reports and community driven peer review. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other servers which we think will help further educate researchers about the benefits of using community driven preprint servers to post their research findings.

@@

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses. Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem. As a result, articles on *EcoEvoRxiv* are more diverse in nature than other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) (@fig-summary).

@@

*EcoEvoRxiv* is the only preprint server that has worked hard to break down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientitic knowledge, Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in our other accepted langauges (e.g., French and Portuguese). Part of the challenge in getting authors to submit multilingual preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries; cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

We were interested in getting an understanding of whether career stage is related to preprint use. We used the first publication year of the submitting author as a proxy for career stage. We predicted that early career researchers would be more open to using preprints in Ecology and Evolution. Interestingly, there was strong evidence that the number of preprints posted decreases with later career stages (negative binomial glm: year slope = `r res[2, "Estimate"]`, SE: `r res[2, "Std. Error"]`, *p* < 0.001, N = `r nrow(count_dat[complete.cases(count_dat),])`). The majority of preprints were submitted by authors who published their first paper in the last 10 years ([@fig-summary]E), with the median year of first publication of submitting authors being `r median(data2$submitting_author_first_publication_year, na.rm = TRUE)` (mean = `r round(mean(data2$submitting_author_first_publication_year, na.rm = TRUE),0)`; SD = `r sd(data2$submitting_author_first_publication_year, na.rm = TRUE)`).

Gender differences in preprint use have been established in a number of other research fields. It is hypothsised that such differences are driven by the larger childcare burden of female researchers, which limit their abilities to publish as frequently as their male counterparts [@ucar2022mind; @wehner2020comparison]. Understanding gender publishing patterns is challenging with observational data. We cannot know the gender of authors for certainty but we can use a data driven approach to ascertain the probability that a particular name is of a given gender (male or female). We used the R package *DemografixeR* (vers. `r utils::packageVersion("DemografixeR")`; @RJ-2016-002) to predict the most likely gender of the submitting author of a preprint. We incorperated information about the submitting authors country of affiliation to better account for differences across countries in naming. We only used names where the gender was identified with 95% certainty.

Overall, females were less likely to publish preprints on *EcoEvoRxiv* compared to males (females: `r (table(data4$gender)[1] / sum(table(data4$gender))*100)`% compared to males: `r (table(data4$gender)[2] / sum(table(data4$gender))*100)`%), which is likely a reflection of the broader publishing disparities between male and feamle scientists.

# Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication, and beyond

```{r, fig-pubsummary}
#| label: fig-pubsummary
#| fig-cap: Summary of the publication status of preprints on *EcoEvoRxiv*.  B) Time between submission and publication of preprints on *EcoEvoRxiv*. 'Preprints' that were published within 2 months (60 days) of being submitted are considered postprints.

@@

In total, `r unpub` papers were unpublished out of the `r nrow(data2)` (`r (unpub / nrow(data2))*100`%). The maximum time between submission and publication was `r sum_preprints$max` days (@fig-pubsummary A). The median time to publication was `r sum_preprints$median` days (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days).

Preprints can be a useful way to ensure that research remains open and accessible to the research community as the paper is available online no matter what the published status of the preprint ends up being. Most of the published preprints ended up being open access(@fig-pubsummary A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The open access status of the papers that were published in open access journals also varied widely (@fig-pubsummary B). Data and code sharing is a key component of open science. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code.

Authors may be reluctant to share data in code for preprints because of the perceived concern that others may acquire and use such data and code without authors' consent. However, while we found that `r pre_article_data[3,2]` (`r pre_article_data[3,"percent"]*100`%) of preprints that did not share data at the preprint stage did in the published article, there were still `r pre_article_data[2,2]` (`r pre_article_data[2,"percent"]*100`%) that never ended up sharing data. The same was true for code. Overall, while `r pre_article_code[3,2]` (`r pre_article_code[3,"percent"]*100`%) of preprints that did not share code at the preprint stage did in the published article, there were still `r pre_article_code[2,2]` (`r pre_article_code[2,"percent"]*100`%) that never ended up sharing code.

@@

- First, authors need not be afraid of sharing data and code early in the publication process. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and postprints. Posting preprints is expected to help: 1) establish precedence of research findings; 2) ensure that research is open access and 3) authors receive feedback on their research before it is published. While it is clear points 1 and 2 have already been achived discussions around a preprints findings are still rare. It is clear that we, as a community, need to provide better platforms that document discussions around a preprint findings. Such discussions are useful for the authors to help improve their work and communicate their findings more effectively (when done constructively, of course).
- Third, authors need to be aware of the benefits of using community driven preprint servers. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers. 
- Lastly, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process.

While most preprints seemlessly get connected with their published version and merged together often they can remain disconnected making it unclear how to associate the preprint and published version, creating confusion in the findings and leading to frustration from authors. Cross-refererencing problems do occur and are the major reason for why preprints are not always connected automatically. Thanksfully, the reasons are fairly well understood and easily rectified. They often result from a mismatch between titles or authors in preprint and published metadata. In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%) and this was the major contributer to preprints and published articles not being matched automatically in Google Scholar during our data collection. As such, we would recommend that authors make sure that, when preprints end up getting published, that they update their preprint with the publication DOI when it's available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged and the published paper is ultimately open access, allowing authors to save open access fees.

>>>>

Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) encouraging constructive and critical feedback from a research community; 3) helping establish precedence of research findings; and 4) removing financial barriers to publication [@Proulx2013; @bourne2017ten; @vale2015accelerating]. *EcoEvoRxiv* is a community driven preprint server that has paved the way for new initiatives, including accepting multilingual preprints, recognising and accepting registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI).

Preprint servers, such as *EcoEvoRxiv*, can empower researchers to make their findings more accessible, open and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives, improve others, and allow us to work harder in further breaking down barriers to their use in our field.

Here, we explore the first preprints uploaded to *EcoEvoRxiv*, one of the few community run preprint servers, to understand preprint practices and use in ecology and evolution. We focus our attention on understanding better: 1) how long it takes for a preprint to become published, and how many preprints remain unpublished; 2) what countries make use of preprints; 3) how career stage and gender impact preprint use; 4) whether data and code are more likely to be shared in preprints; and 5) the extent to which authors make use of preprint servers for registered reports and community driven peer review. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other servers which we think will help further educate researchers about the benefits of using community driven preprint servers to post their research findings.

@@

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses. Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem [@haddaway2015shades; @haddaway2020eight]. As a result, articles on *EcoEvoRxiv* are more diverse in nature than other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) (@fig-summary).

@@

A significant barrier to the communication of research findings is due to the fact that they are primarily communicated in English [@amano2016languages; @amano2013four; @amano2021ten]. Communication of research through a single language has major consequences for the global distribution of knowledge resulting in knowledge gaps in some of the most biodiverse and threatened regions in the world [@amano2023role; @zenni2023multi]. Such gaps also have flow on effects for research syntheses and meta-analyses because it creates a distorted picture of our knowledge base that can lead to bias and inadequate sampling that affects policy development and decision making [@hannah2024language; @zenni2023multi; @white2021geographical; @konno2020ignoring].

*EcoEvoRxiv* is the only preprint server breaking down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientitic knowledge. Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in other accepted langauges (e.g., French and Portuguese). Part of the challenge in getting authors to submit multilingual preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries and the fact that it accepts preprints in different languages, cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

Research papers take time to be published (see below). Early and mid career researchers (ECMRs) need their research to come out quickly for job applications, promotion and obtaining grants to accerate their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are one way ECMRs can acheive faster dissemination and greater visibility. As such, we expect ECMRs to accept and adopt preprints more than later career colleagues. At the same time, gender differences in preprint use and publication outcomes  have been established in a number of other research fields, including in ecology and evolution [@fox2019gender; @wehner2020comparison]. We therefore expect such discrepencies to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certainty but we can use a data driven approach to ascertain the probability that a particular name is of a given gender (male or female). To obtain an idea of an authors' gender, we used the R package *DemografixeR* (vers. `r utils::packageVersion("DemografixeR")`; @RJ-2016-002) to predict the most likely gender of the submitting author of a preprint. We incorperated information about the submitting authors country of affiliation to better account for differences across countries. We only used names where the gender was identified with 95% certainty.

Overall, females were less likely to publish preprints on *EcoEvoRxiv* compared to males (females: `r (table(data4$gender)[1] / sum(table(data4$gender))*100)`% compared to males: `r (table(data4$gender)[2] / sum(table(data4$gender))*100)`%), which is likely a reflection of the broader publishing disparities between male and female scientists [@fox2019gender]. There was also strong evidence that the number of preprints posted decreases with later career stages (negative binomial glm: year slope = `r res[2, "Estimate"]`, SE: `r res[2, "Std. Error"]`, *p* < 0.001, N = `r nrow(count_dat[complete.cases(count_dat),])`). The majority of preprints were submitted by authors who published their first paper in the last 10 years ([@fig-summary]E), with the median year of first publication of submitting authors being `r median(data2$submitting_author_first_publication_year, na.rm = TRUE)` (mean = `r round(mean(data2$submitting_author_first_publication_year, na.rm = TRUE),0)`; SD = `r sd(data2$submitting_author_first_publication_year, na.rm = TRUE)`). These preprinting practices support the idea that ECMRs may value preprints as a way to make their work more visible and disseminate tehir finding more quickly.

# Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| fig-cap: Summary of the publication status of preprints on *EcoEvoRxiv*.  A) Time between submission and publication of preprints on *EcoEvoRxiv*. 'Preprints' that were published within 2 months (60 days) of being submitted are considered postprints. B) Number of preprints that were published in open access journals. C) Open access status of preprints published on *EcoEvoRxiv*.

@@

### **Science takes time, but publication takes longer**
Increased competition in science has resulted in the bar being raised with respect to the amount of data required for publication [@vale2015accelerating]. This is a good outcome if it results in higher impact research that better clarifies our understadning of the natural world, but it does come at a cost for the rate at which research findings can be disseminated [@vale2015accelerating]. Preprinting has been proposed as a way research can be disseminated more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is sparse, and the real benefit of preprints in this context would be reinforced if preprinting is able to dramatically speed up dissemination. We evaluated just how much time it takes to publish a research paper in ecology and evolution by ascertaining the time between when a preprint was first posted to when it was finally accepted in a research journal. In total, `r unpub` papers were still unpublished out of the `r nrow(data2)` (`r (unpub / nrow(data2))*100`%) at the time these data were collected. Astonishingly, the maximum time between submission and publication was `r sum_preprints$max` days ([@fig-pubsummary]A), and the median time to publication was `r sum_preprints$median` days (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days). Our results largely confirm the long timespans most authors are dealing with to get their research published after the paper is written.

### **Catious 'open'-mindedness of research findings in preprints**
In additon to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using....LOSIA?. Most of the published preprints ended up being open access([@fig-pubsummary]A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The level of open access for a paper published in an open access journal also varied widely ([@fig-pubsummary]B).

Data and code sharing is also a key component of open science [@roche2015public]. In the spirit of 'openness' we expected data and code sharing among preprints to be better than many papers published in research journals [@roche2015public; @o2021preferred]. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code. Authors may be reluctant to share data in code for preprints because of the perceived concern that others may acquire and use such data and code without authors' consent. However, while `r pre_article_data[3,2]` (`r pre_article_data[3,"percent"]*100`%) of preprints that did not share data at the preprint stage did ultimately share data in the published article, there were still `r pre_article_data[2,2]` (`r pre_article_data[2,"percent"]*100`%) that never ended up sharing data. The same was true for code. Overall, `r pre_article_code[3,2]` (`r pre_article_code[3,"percent"]*100`%) of preprints did not share code at the preprint stage but did in the published article, however, there were still `r pre_article_code[2,2]` (`r pre_article_code[2,"percent"]*100`%) that never ended up sharing code. Code and data sharing patterns in our sample of preprints is consistent with other analyses looking at code and data sharing in published articles (e.g., [@o2021preferred]), even for journals that have strict public data archiving policies [@roche2015public].

@@

- First, authors need not be afraid of sharing data and code early in the publication process. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and postprints. Posting preprints is expected to help: 1) establish precedence of research findings; 2) ensure that research is open access and 3) authors receive feedback on their research before it is published [@bourne2017ten]. While it is clear points 1 and 2 have already been achived discussions around a preprints findings are still rare, which is consistent with findings on other preprint servers (e.g. BioRxiv [@anderson2020biorxiv]). It is clear that we, as a community, need to provide better platforms that document discussions around a preprint findings. Such discussions are useful for the authors to help improve their work and communicate their findings more effectively (when done constructively, of course).
- Third, authors need to be aware of the benefits of using community driven preprint servers and the unique initiatives being pursued. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers and those volunteering their time at *EcoEvoRxiv* remain open to new and innovative ways to improve publication and open science practices. 
- Lastly, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process.

While most preprints get seemlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion in research findings and lead to frustration from authors. Thankfully, the reasons for unmatched preprints and publications are fairly well understood and easily rectified. They often result from a mismatch between titles or authors in preprint and published metadata. In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). Mismatched metadata was the almost always the contributer to preprints and published articles not being matched automatically in Google Scholar during our data collection. As such, we would recommend that authors make sure that, when preprints end up getting published, that they update their preprint with the publication DOI when it is available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the reserach findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.
